\large{\bf Looking forwards and backwards through locally regulated populations }

### 2023-03-11 forwards-and-backwards.tex
good citation
rewrite with gamma B1 assumption

\comment{TO CITE: \citet{ghosh2022emergent}}

@@

Most work on this topic has focused on models that caricature the classical Fisher-KPP equation with a 
stochastic term, i.e.
$$dw=\big(\Delta w +sw(1-w)\big)dt +\sqrt{\frac{\alpha(w)}{N}}W(dt,dx),$$
where $W$ is space-time white noise, and $N$ is a measure of the 
local population density. The coefficient
$\alpha(w)$ is generally taken to be either
$w$, corresponding to a superprocess limit, or $w(1-w)$ giving a 
spatial analogue of a Wright-Fisher diffusion. 
For very large $N$, and on suitable timescales, 
starting with the pioneering work of~\cite{brunet/derrida/mueller/munier:2006},
a considerable body of evidence has been amassed to underpin the
conjecture that for this, and a wide class of 
related models, 
genealogies converge to a 
Bolthausen-Sznitman coalescent. %, e.g.~\cite{brunet/derrida/mueller/munier:2007}.
This reflects the fact that, for this equation, 
ancestral lineages become trapped in 
the wavefront, where the growth rate of the population is highest. 
Once there, they will experience 
rapid periods of coalescence 
corresponding to significant proportions of individuals in the front being 
descended from particularly reproductively successful ancestors.

@@

We assume that 
\begin{enumerate}
\item The function $F(x,m)$ satisfies 
\begin{enumerate}
\item $F(x,m)$ is locally Lipschitz in $m$;
\item $F(x,m)$ is uniformly bounded above (but not necessarily below);
\item for each fixed $m$, $\sup_{x\in\IR^d}\sup_{k\leq m}|F(x,k)|<\infty$;
\end{enumerate}
\item The functions $r(x,m)$, $\gamma(x,m)$ have bounded first and second 
derivatives in both arguments;
\item $\gamma(x,m)$ is uniformly bounded;
\item $|\gamma(x, \smooth{\gamma} \eta(x)) \theta \int (r(y, \smooth{r} \eta(y)) - r(x, \smooth{r} \eta(x))) q_\theta(x, dy)|$
    is uniformly bounded for $x \in \IR^d$ and $\eta \in \measures$.
\end{enumerate}

@@

Comparing %\eqref{PME} 
with \eqref{eqn:nonlocal_pde},
we see that to set up a limit in which the population density $\varphi$ follows the 
porous medium equation with logistic growth of~(\ref{PME}),
we need $r=1$,
$\gamma = \varphi$, and $F = 1 - \varphi$.
Consulting equation~\eqref{eqn:mu_defn},
this implies that $\mu_\theta = \max\big(0, (1 + 1/\theta) \varphi - 1/\theta\big)$.
In other words,
establishment is certain
and birth rates increase linearly with population density,
but to compensate, death rates increase slightly faster (also linearly),
as shown in Figure~\ref{fig:pme_waves}.
Alert readers will notice that
the condition from Assumptions~\ref{def:model_setup} 
that $\gamma (x,m)$ be uniformly bounded 
is violated.
This can be corrected by use of a cut-off, and in fact the downwards drift
provided by the logistic control of the population size prevents $m$ from 
getting too big. 
In practice in our simulations we choose a suitably small value for $\beta$, define
\begin{align*}
    \gamma(x, m) = (1 - \exp(- \beta m)) / \beta, \qquad
    F(x, m) = \exp(- \beta m) / \beta ,
\end{align*}
and set $\mu_\theta(x, m) = \max(0, \gamma - F / \theta)$.
\comment{I don't understand this - something has gone wrong.}
\comment{TODO: peter to look into this}
Birth and death rates are equal at density $m = 1$,
corresponding to an unscaled density of $N$ individuals per unit area.

@@

\comment{In this section I haven't specified the filtration for the martingale problem.}

@@

\comment{the next theorem needs rewriting along the lines of Thm A.12 in Kurtz and Rodrigues.}
\comment{TODO: it's been rewritten some; but needs checking against that Thm yet}

@@

@@

\begin{proof}[of Lemma~\ref{lem:conditions on r}]
    TODO MOVE PROOF HERE
\end{proof}

@@

\begin{proof}
    Consider the semimartingale decomposition from equation~\eqref{eqn:eta_martingale}:
    \begin{align} \label{eqn:eta_f_mgale_decomp}
        \langle f, \eta^N_t \rangle
        &=
        \langle f, \eta^N_0 \rangle
        + \int_0^t \int_{\IR^d} \big\{
            \gamma(x, \eta^N_s) B_f(x, \eta^N_s)
            + f(x) F(x, \eta^N_s)
            \big\} \eta^N_s(dx) ds
        + M^N_t(f) ,
    \end{align}
    where $M^N_t(f)$ is a martingale and, for the moment, we are defining
    \begin{equation}
\label{generator Bf}
        B_f(x, \eta) = \int_{\IR^d} \theta (f(y) r(y, \eta) - f(x) r(x, \eta)) q_\theta(x, dy) .
    \end{equation}

\comment{TODO MOVE SOME OF THIS TO PROOF OF LEMMA~\ref{lem:conditions on r}}
We need a bound on $\gamma(x,\eta)B_f(x,\eta)+F$.
First suppose that
Condition~\ref{control through r} of  
Assumptions~\ref{def:model_setup} is satisfied.
We write
    \begin{align*}
        f(y)r\big(y,\rho_r*\eta(y)\big) - f(x)r\big(x,\rho_r*\eta(x)\big)
        &=
        \sum_i (y_i - x_i) \partial_{x_i} \Big(f(x)r\big(x,\rho_r*\eta(x)\big)\Big) 
        \\ &\qquad {}
        +
        \frac{1}{2} \sum_{ij} (y_i - x_i)(y_j - x_j)
            \partial_{x_i x_j} \Big(f(z)r\big(\smooth{r}\eta(z)\big)\Big) ,
    \end{align*}
    where $z = \kappa x + (1-\kappa) y$ for some $0 \le \kappa \le 1$. Condition~\ref{control through r} 
gives uniform bounds on the derivatives of $r(x, \rho_r*\eta(x))$ 
in this expression and so, provided $f$ also has uniformly bounded first
and second derivatives, we have a bound of the form
\begin{multline*}
\Big| f(y)r\big(y,\rho_r*\eta(y)\big) - f(x)r\big(x,\rho_r*\eta(x)\big) \Big|
\\
\leq
        K_1 \left|
            \theta \int_{\IR^d} \sum_i (y_i - x_i) q_\theta(x, dy)
        \right|
        +
        K_2 \left|
            \theta \int_{\IR^d} \sum_{ij} (y_i - x_i) (y_j - x_j)  q_\theta(x, dy)
        \right| 
\end{multline*}
for suitable constants $K_1$, $K_2$.
This is bounded uniformly because
$q_\theta(x, dy)$ is the density of a Gaussian with mean $\meanq(x)/\theta$
and covariance $\covq(x)/\theta$, and both $\meanq(x)$ and $\covq(x)$
are bounded uniformly in $x$ under Assumptions~\ref{def:model_setup}.

Now suppose that instead
Condition~\ref{control through gamma} of Assumptions~\ref{def:model_setup}
is satisfied.
First note that,  
    \begin{align}
	    \nonumber
    &
        \left| \theta 
            \int_{\IR^n} 
                \big\{
                    f(y) r\big(y,\smooth{r}\eta(y)\big)
                    -
                    f(x) r\big(x, \smooth{r}\eta(x)\big)
                \big\}
            q_{\theta}(x,dy)
        \right|
    \\ 
 \label{eqn:q_bound_split}
	    &\qquad {} \leq
        \left| \theta 
            \int_{\IR^n} 
                \big\{
                    f(y) r\big(y, \smooth{r}\eta(y)\big)
                    -
                    f(x) r\big(x,\smooth{r}\eta(y)\big)
                \big\}
            q_{\theta}(x,dy)
        \right|
    \\ 
\nonumber
	    &\qquad \qquad {}  +
        \left| \theta 
            \int_{\IR^n} 
                \big\{
                    f(x) r\big(x,\smooth{r}\eta(y)\big)
                    -
                    f(x) r\big(x,\smooth{r}\eta(x)\big)
                \big\}
            q_{\theta}(x,dy) 
        \right| .
    \end{align}

 Writing
    $K_3 = \sup_{x,m} \max_i |\partial_{x_i} f(x)r(x, m)|$
    and 
    $K_4 = \sup_{x,m} \max_{i,j} |\partial_{x_i x_j} f(x)r(x, m)|$, since
\[\big|\partial_{x_i}\big(f(x)r(x,\rho_r*\eta(y))\big|\leq K_3\]
and
\[\big|\partial_{x_ix_j}\big(f(x)r(x,\rho_r*\eta(y))\big|\leq K_4,\]
the first term
is bounded exactly as above. For the second, 
    \begin{multline*}
        r(x, \smooth{r}\eta(y)) - r(x, \smooth{r}\eta(x))
        \\
=
        (\smooth{r}\eta(y) - \smooth{r}\eta(x)) r'(x,\smooth{r}\eta(x))
        + \frac{1}{2} (\smooth{r}\eta(y) - \smooth{r}\eta(x))^2 r''(x, \overline{m}),
    \end{multline*}
    where $\overline{m}= \kappa'\smooth{r}\eta(x) +(1-\kappa') 
\smooth{r}\eta(y)$ for some $0\leq \kappa'\leq 1$, and
    we have used $r'$ and $r''$ to denote the first and second derivatives of $r(x,m)$
with respect to the second argument.
    So, writing $K_5=\|f\|_\infty\|r'\|_\infty$ and $K_6=\|f\|_\infty\|r''\|_\infty$, in
this case 
    the term in~\eqref{eqn:q_bound_split} is bounded by
    \begin{align*}
       K_5
        \left|
            \theta \int_{\IR^d}
                (\smooth{r}\eta(y) - \smooth{r}\eta(x))
            q_\theta(x, dy)
        \right|
        +
       K_6
        \left|
            \theta \int_{\IR^d}
                (\smooth{r}\eta(y) - \smooth{r}\eta(x))^2
            q_\theta(x, dy)
        \right| .
    \end{align*}
Under Condition~\ref{control through gamma} of 
Assumptions~\ref{def:model_setup}, the right hand side is bounded by 
a constant times $\smooth{\gamma}\eta(x)+(\smooth{\gamma}\eta(x))^2$ and 
$m^2\gamma(m)$ is bounded. We have shown that $\gamma B_f$ is bounded, and so recalling that
$F$ is bounded above, we can now conclude that
under Assumptions~\ref{def:model_setup}
    $\gamma(x, \eta) B_f(x, \eta) + f(x) F(x, \eta) \le C_f $ for some $C_f$.

>>>>

@@

Most work on this topic has focused on models that caricature the classical Fisher-KPP equation with a 
stochastic term, i.e.
$$dw=\big(\Delta w +sw(1-w)\big)dt +\sqrt{\frac{\alpha(w)}{N}}W(dt,dx),$$
where $W$ is space-time white noise, and $N$ is a measure of the 
local population density. The coefficient
$\alpha(w)$ is generally taken to be either
$w$, corresponding to a superprocess limit, or $w(1-w)$ giving a 
spatial analogue of a Wright-Fisher diffusion. 
For instance, \citep{ghosh2022emergent}
derives a stochastic version of~\eqref{semilinear PDE} to describe abundances of mutant bacteria strains
along the human gut.
For very large $N$, and on suitable timescales, 
starting with the pioneering work of~\cite{brunet/derrida/mueller/munier:2006},
a considerable body of evidence has been amassed to underpin the
conjecture that for this, and a wide class of 
related models, 
genealogies converge to a 
Bolthausen-Sznitman coalescent. %, e.g.~\cite{brunet/derrida/mueller/munier:2007}.
This reflects the fact that, for this equation, 
ancestral lineages become trapped in 
the wavefront, where the growth rate of the population is highest. 
Once there, they will experience 
rapid periods of coalescence 
corresponding to significant proportions of individuals in the front being 
descended from particularly reproductively successful ancestors.

@@

We assume that 
\begin{enumerate}
\item The function $F(x,m)$ satisfies 
    \begin{enumerate}
    \item $F(x,m)$ is locally Lipschitz in $m$;
    \item $F(x,m)$ is uniformly bounded above (but not necessarily below);
    \item for each fixed $m$, $\sup_{x\in\IR^d}\sup_{k\leq m}|F(x,k)|<\infty$;
    \end{enumerate}
\item The functions $r(x,m)$, $\gamma(x,m)$ have bounded first and second 
derivatives in both arguments;
\item $\gamma(x,m)$ is uniformly bounded;
\item $|\gamma(x, \smooth{\gamma} \eta(x)) \theta \int (r(y, \smooth{r} \eta(y)) - r(x, \smooth{r} \eta(x))) q_\theta(x, dy)|$
    is uniformly bounded for $x \in \IR^d$ and $\eta \in \measures$.
        \label{gamma_B_condition}
\end{enumerate}

@@

Comparing %\eqref{PME} 
with \eqref{eqn:nonlocal_pde},
we see that to set up a limit in which the population density $\varphi$ follows the 
porous medium equation with logistic growth of~(\ref{PME}),
we need $r=1$,
$\gamma = \varphi$, and $F = 1 - \varphi$.
Consulting equation~\eqref{eqn:mu_defn},
this implies that $\mu_\theta = \max\big(0, (1 + 1/\theta) \varphi - 1/\theta\big)$.
In other words,
establishment is certain
and birth rates increase linearly with population density,
but to compensate, death rates increase slightly faster (also linearly),
as shown in Figure~\ref{fig:pme_waves}.
Alert readers will notice that
the condition from Assumptions~\ref{def:model_setup} 
that $\gamma(x,m)$ be uniformly bounded 
is violated.
This can be corrected by use of a cut-off, and in fact the downwards drift
provided by the logistic control of the population size prevents $m$ from 
getting too big. 
In practice in our simulations we
take discrete time steps of length $dt$ (with $dt$ suitably small),
and have each individual with local density $m$ reproduce and die, respectively, with probabilities
\begin{align*}
    p_\text{birth}(m) = \left(1 - e^{- m dt}\right)
    \qquad 
    p_\text{death}(m) = \left(1 - e^{- (m (1+1/\theta) - 1/\theta) dt} \right) ,
\end{align*}
so that $\gamma(x, m) = p_\text{birth}(m)/dt \approx m$
and
\begin{align*}
    F(x, m) = \theta(\gamma(x, m) - \mu(x, m)) / dt \approx 1 - m .
\end{align*}
Birth and death rates are equal at density $m = 1$,
corresponding to an unscaled density of $N$ individuals per unit area.

@@

@@

@@

Below we will have frequent use for the quantity
\begin{equation}
\label{generator Bf}
    B_f(x, \eta) = \theta \int_{\IR^d} (f(y) r(y, \eta) - f(x) r(x, \eta)) q_\theta(x, dy) .
\end{equation}

@@

\begin{proof}[of Lemma~\ref{lem:conditions on r}]
Here, we need to prove that $|\gamma(x, \eta) B_1(x, \eta)|$ is bounded,
uniformly over $x$ and $\eta$.
First suppose that
Condition~\ref{control through r} of  
Lemma~\ref{lem:conditions on r} is satisfied.
We write
\begin{align*}
    r\big(y,\rho_r*\eta(y)\big) - r\big(x,\rho_r*\eta(x)\big)
    &=
    \sum_i (y_i - x_i) \partial_{x_i} \Big(r\big(x,\rho_r*\eta(x)\big)\Big) 
    \\ &\qquad {}
    +
    \frac{1}{2} \sum_{ij} (y_i - x_i)(y_j - x_j)
        \partial_{x_i x_j} \Big(r\big(\smooth{r}\eta(z)\big)\Big) ,
\end{align*}
where $z = \kappa x + (1-\kappa) y$ for some $0 \le \kappa \le 1$.
Condition~\ref{control through r} 
gives uniform bounds on the derivatives of $r(x, \rho_r*\eta(x))$ 
in this expression and so we have a bound of the form
\begin{multline*}
\Big| r\big(y,\rho_r*\eta(y)\big) - r\big(x,\rho_r*\eta(x)\big) \Big|
\\
\leq
        K_1 \left|
            \theta \int_{\IR^d} \sum_i (y_i - x_i) q_\theta(x, dy)
        \right|
        +
        K_2 \left|
            \theta \int_{\IR^d} \sum_{ij} (y_i - x_i) (y_j - x_j)  q_\theta(x, dy)
        \right| 
\end{multline*}
for suitable constants $K_1$, $K_2$.
This is bounded uniformly because
$q_\theta(x, dy)$ is the density of a Gaussian with mean $\meanq(x)/\theta$
and covariance $\covq(x)/\theta$, and both $\meanq(x)$ and $\covq(x)$
are bounded uniformly in $x$ under Assumptions~\ref{def:model_setup}.
u implies that $|B_1|$ is uniformly bounded;
since $\gamma$ is bounded as well, so is $|\gamma B_1|$.

Now suppose that instead
Condition~\ref{control through gamma} of Lemma~\ref{lem:conditions on r} is satisfied.
is satisfied.
First note that
\begin{multline*}
    r(x, \smooth{r}\eta(y)) - r(x, \smooth{r}\eta(x))
    \\
=
    (\smooth{r}\eta(y) - \smooth{r}\eta(x)) r'(x,\smooth{r}\eta(x))
    + \frac{1}{2} (\smooth{r}\eta(y) - \smooth{r}\eta(x))^2 r''(x, \overline{m}),
\end{multline*}
where $\overline{m}= \kappa'\smooth{r}\eta(x) +(1-\kappa') 
\smooth{r}\eta(y)$ for some $0\leq \kappa'\leq 1$, and
we have used $r'$ and $r''$ to denote the first and second derivatives of $r(x,m)$
with respect to the second argument.
So, in this case
\begin{align*}
&
\left| \theta 
    \int_{\IR^n} 
        \big\{
            r\big(x,\smooth{r}\eta(y)\big)
            -
            r\big(x,\smooth{r}\eta(x)\big)
        \big\}
    q_{\theta}(x,dy) 
\right| 
\\ &\qquad
\le
   \|r'\|_\infty
    \left|
        \theta \int_{\IR^d}
            (\smooth{r}\eta(y) - \smooth{r}\eta(x))
        q_\theta(x, dy)
    \right|
    +
   \|r''\|_\infty
    \left|
        \theta \int_{\IR^d}
            (\smooth{r}\eta(y) - \smooth{r}\eta(x))^2
        q_\theta(x, dy)
    \right| .
\end{align*}
Therefore,
under Condition~\ref{control through gamma}, $|B_1(x,\eta)|$ is bounded by 
a constant times $\smooth{\gamma}\eta(x)+(\smooth{\gamma}\eta(x))^2$;
since $m^2\gamma(m)$ is bounded, so is $|\gamma(x,\eta) B_1(x,\eta)$.
\end{proof}

@@

\begin{proof}
Consider the semimartingale decomposition from equation~\eqref{eqn:eta_martingale}:
\begin{align} \label{eqn:eta_f_mgale_decomp}
    \langle f, \eta^N_t \rangle
    &=
    \langle f, \eta^N_0 \rangle
    + \int_0^t \int_{\IR^d} \big\{
        \gamma(x, \eta^N_s) B_f(x, \eta^N_s)
        + f(x) F(x, \eta^N_s)
        \big\} \eta^N_s(dx) ds
    + M^N_t(f) ,
\end{align}
where $M^N_t(f)$ is a martingale.
% and we are defining
% \begin{equation}
% \label{generator Bf}
%     B_f(x, \eta) = \int_{\IR^d} \theta (f(y) r(y, \eta) - f(x) r(x, \eta)) q_\theta(x, dy) .
% \end{equation}

We need a bound on $\gamma(x,\eta)B_f(x,\eta)+fF$.
First note that we may rewrite $B_f(x, \eta)$ as
\begin{align}
    \nonumber
&
    \left| \theta 
        \int_{\IR^n} 
            \big\{
                f(y) r\big(y,\smooth{r}\eta(y)\big)
                -
                f(x) r\big(x, \smooth{r}\eta(x)\big)
            \big\}
        q_{\theta}(x,dy)
    \right|
\\ 
    &\qquad {} =
    \Big| \theta 
        \int_{\IR^n} 
            ( f(y) - f(x) )
            r\big(y, \smooth{r}\eta(y)\big)
        q_{\theta}(x,dy)
\\ 
\nonumber
    &\qquad \qquad {}  +
    \theta 
            f(x)
        \int_{\IR^n} 
            \big\{
                r\big(y,\smooth{r}\eta(y)\big)
                -
                r\big(x,\smooth{r}\eta(x)\big)
            \big\}
        q_{\theta}(x,dy) 
    \Big| 
\\ 
    &\qquad {} \leq
    \left| \theta 
        \int_{\IR^n} 
            ( f(y) - f(x) )
        q_{\theta}(x,dy)
    \right|
    +
    \|f\|_\infty |B_1(x, \eta)| .
\end{align}
The first term is bounded uniformly
because $q_\theta(x, dy)$ is the density of a Gaussian with mean $\meanq(x)/\theta$
and covariance $\covq(x)/\theta$,
and both $\meanq(x)$ and $\covq(x)$
are bounded uniformly in $x$ under Assumptions~\ref{def:model_setup}.
Furthermore, Condition~\ref{gamma_B_condition} of Assumptions~\ref{def:model_setup}
stipulates that $|\gamma(x, \eta) B_1(x, \eta)|$ is uniformly bounded,
and so recalling that $F$ is bounded above, we can now conclude that
under Assumptions~\ref{def:model_setup}
$\gamma(x, \eta) B_f(x, \eta) + f(x) F(x, \eta) \le C_f $ for some $C_f$.
